---
title: "Lending Club"
format: html
editor: source
---

## Loading the data

```{r}
library(tidymodels)
library(modeldata)
head(lending_club)
```

```{r}
library(tidymodels)
library(modeldata)

# 1. Load Data
data("lending_club")

# 2. Create Train/Test Sets
set.seed(123)
lc_split <- initial_split(lending_club, prop = 0.80, strata = funded_amnt)
lc_train <- training(lc_split)
lc_test  <- testing(lc_split)

# 3. Create a Recipe (Preprocessing)
# We handle categorical variables and remove zero-variance predictors
lc_rec <- recipe(funded_amnt ~ ., data = lc_train) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())

# 4. Define 10-fold Cross-Validation
lc_folds <- vfold_cv(lc_train, v = 10, strata = funded_amnt)


### Model Specifications (Parsnip)

# Model 1: Linear Regression (Penalty tuning)
lm_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# Model 2: Random Forest (Mtry tuning)
rf_spec <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Model 3: XGBoost (Learning rate and Tree depth tuning)
xgb_spec <- boost_tree(tree_depth = tune(), learn_rate = tune(), loss_reduction = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

### Workflow and Tuning (Example for Linear Regression)

# Create Workflow
lc_workflow <- workflow() %>%
  add_recipe(lc_rec) %>%
  add_model(lm_spec)

# Tune Model
tuning_results <- tune_grid(
  lc_workflow,
  resamples = lc_folds,
  grid = 10
)

# 5. Finalize Model and Evaluate on Test Set
best_params <- select_best(tuning_results, metric = "rmse")
final_wf <- finalize_workflow(lc_workflow, best_params)
final_fit <- last_fit(final_wf, lc_split)

# 6. Report Test RMSE and R-squared
collect_metrics(final_fit)
```


## Create `train` and `test` sets.

```{r}
set.seed(123)
lc_split <- initial_split(lending_club, 
                          prop = 0.80, 
                          strata = funded_amnt)
lc_train <- training(lc_split)
lc_test  <- testing(lc_split)
```

## Create a Recipe (Preprocessing)

```{r}
# Handle categorical variables and remove zero-variance predictors
lc_rec <- recipe(funded_amnt ~ ., data = lc_train) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors())
```

## Define 10-fold Cross-Validation

```{r}
lc_folds <- vfold_cv(lc_train, v = 3, strata = funded_amnt)
```

## Create Models

```{r}
# Model 1: Elastic Net Regression
lm_spec <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("regression")

# Model 2: Random Forest
rf_spec <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) |>
  set_engine("ranger") |>
  set_mode("regression")

# Model 3: XGBoost
xgb_spec <- boost_tree(tree_depth = tune(), learn_rate = tune(), 
                       loss_reduction = tune(), stop_iter = 10) |>
  set_engine("xgboost") |>
  set_mode("regression")
```

## Workflow and Tuning

`workflow_set`: Instead of writing separate tuning loops, this bundles your Linear, RF, and XGBoost models together to run them simultaneously.

**Hyperparameter Grids**: It uses `grid = 10` to automatically pick 10 sensible candidate values for parameters like `mtry`, `learn_rate`, and `penalty`.

Evaluation: It compares them using the 10-fold cross-validation required by Project One Directions and outputs the final metrics.

```{r}
# Combine all models into a set
lc_models <- workflow_set(
  preproc = list(simple = lc_rec),
  models = list(linear = lm_spec, rf = rf_spec, xgboost = xgb_spec)
)

# Run tuning across the grid for all models
# Note: 'grid = 10' tries 10 random combinations per model
grid_results <- lc_models |>
  workflow_map("tune_grid", resamples = lc_folds, grid = 3, metrics = metric_set(rmse, rsq))
```

## Finalize Model and Evaluate on Test Set

```{r}
autoplot(grid_results) # Visualizes RMSE vs R-squared across models
best_result <- grid_results |> 
  extract_workflow_set_result("simple_xgboost") |> 
  select_best(metric = "rmse")

# Final Fit to the Test Set
final_fit <- grid_results |>
  extract_workflow("simple_xgboost") |>
  finalize_workflow(best_result) |>
  last_fit(split = lc_split)

# Display final Test RMSE and R-squared
collect_metrics(final_fit)
```





