---
title: "Class Example"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::pdf_document2
toc: FALSE
---

```{r label = "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE, fig.align = "center")
library(tidyverse)
library(moderndive)
```

# Code chunks (hidden)

```{r label = "g1", echo = FALSE, fig.cap = "Scatterplot of Log10 of Price versus Number of Bedrooms", fig.width = 2.5, fig.height = 2.5}
p1 <- ggplot(data = house_prices, aes(x = bedrooms, y = log10(price))) + 
  geom_point(alpha = 0.05) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  labs(x = "Number of Bedrooms")
p1
```

## Data munging

```{r, label = "f1"}
# removing outlier
house_prices %>% 
  filter(bedrooms < 33) -> house_prices
```

## Scatterplot without and with outlier removed

```{r, label = "b1", fig.cap = "Scatterplots with and without outlier for Log10 Price versus Number of Bedrooms", fig.width = 10, fig.height = 5}
p2 <- ggplot(data = house_prices, aes(x = bedrooms, y = log10(price))) + 
  geom_point(alpha = 0.05) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() + 
  labs(x = "Number of Bedrooms")
library(patchwork)
p1 + p2
```

The left graph in Figure \@ref(fig:b1) contains all of the data in `house_prices` while the right graph in Figure \@ref(fig:b1) removes the outlier in Figure \@ref(fig:g1) and creates a scatterplot without the "outlier".

# Creating models and showing regression output

```{r, label = "m1"}
mod1 <- lm(log10(price) ~ log10(sqft_living) + bedrooms, data = house_prices)
get_regression_table(mod1) -> T1
T1
```
Note: the output from `T1` could be made to look better.  Consider the `kable()` function from `knitr`:

```{r}
knitr::kable(T1, caption = "A different formatting of T1")
```



## Writing the regression equation with \LaTeX and inline R code

The least squares regression equation from regressing `log10` of `price` onto `log10` of `sqft\_living` and `bedrooms` is written in Equation \@ref(eq:lre). 

\begin{equation}
\widehat{\text{log10(price)}} = `r T1$estimate[1]` + `r T1$estimate[2]` \cdot \text{log10(sqft\_living)}  `r T1$estimate[3]` \cdot \text{bedrooms}
(\#eq:lre)
\end{equation}
