---
title: "Practice"
author: "Your Name"
date: 'Last updated: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
library(tidyverse)
```

# Model

Load the `PASWR2` package and use the `VIT2005` data set to create a linear model named `webe` that regresses `totalprice` onto `area` and `age`.

```{r}
# Your Code here
```

## Model Matrix

Create the model matrix $X$ from the model `webe` and use `dim()` to show the dimmensions of $X$.

```{r}
# Your Code here

```

-------------

Multiply the transpose of $X$ by $X$ and store the result in an object named `XTX`.

```{r}
# Your Code here

```

----------------

Store the values of `totalprice` in a $218 \times 1$ matrix named `Y`.

```{r}
# Your Code here

```

---------------

Use the normal equations ($\mathbf{\hat{\beta}} = (\mathbf{X'X})^{-1}\mathbf{X'Y}$) to compute the least squares estimates of $\mathbf{\beta}$.  Store the results in an object named `betahat`.

```{r}
# Your Code here

```

----------------

Extract the MSE from the object `webe`.

```{r}
# Your Code here

```

------------

Compute the estimated variance-covariance matrix ($\hat{\sigma}^2_{\hat{\beta}} = MSE(\mathbf{X'X})^{-1}$) and store the results in an object named `VCM`.

```{r}
# Your Code here

```

--------------

Construct 98% confidence intervals for $\mathbf{\beta}$.

```{r}
# Your Code here

```

------------------

Compute the mean response ($E(Y_h)$) for an apartment that had an area of 100 square meters and was 25 years old in 2005.

```{r}
# Your Code here

```
-----------

Find the standard error for the expected response given $X_h = (1, 100, 25)$.

```{r}
# Your Code here

```

---------

Construct a 90% confidence interval for the mean response given $X_h = (1, 100, 25)$.

```{r}
# Your Code here

```

----------

