---
title: "Practice"
author: "Partial Answers"
date: 'Last updated: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
library(tidyverse)
```

# Model

Load the `PASWR2` package and use the `VIT2005` data set to create a linear model named `webe` that regresses `totalprice` onto `area` and `age`.

```{r}
library(PASWR2)
webe <- lm(totalprice ~ area + age, data = VIT2005)
summary(webe)
```

## Model Matrix

Create the model matrix $X$ from the model `webe` and use `dim()` to show the dimmensions of $X$.

```{r}
X <- model.matrix(webe)
X <- as.matrix(X)
head(X)
dim(X)
```
-------------

Multiply the transpose of $X$ by $X$ and store the result in an object named `XTX`.

```{r}
XTX <- t(X)%*%X
dim(XTX)
XTX
```
----------------

Store the values of `totalprice` in a $218 \times 1$ matrix named `Y`.

```{r}
Y <- VIT2005 %>% 
  select(totalprice) %>% 
  as.matrix()
dim(Y)
```
---------------

Use the normal equations ($\mathbf{\hat{\beta}} = (\mathbf{X'X})^{-1}\mathbf{X'Y}$) to compute the least squares estimates of $\mathbf{\beta}$.  Store the results in an object named `betahat`.

```{r}
XTXI <- solve(XTX)
XTY <- t(X)%*%Y
betahat <- XTXI %*% XTY
betahat
```

----------------

Extract the MSE from the object `webe`.

```{r}
anova(webe)
anova(webe)[3, 3] -> MSE
MSE
# or
summary(webe)$sigma^2 -> MSE2
MSE2
```

------------

Compute the estimated variance-covariance matrix ($\hat{\sigma}^2_{\hat{\beta}} = MSE(\mathbf{X'X})^{-1}$) and store the results in an object named `VCM`.

```{r}
VCM <- MSE*XTXI
VCM
```

--------------

Construct 98% confidence intervals for $\mathbf{\beta}$.

```{r}
CT <- qt(.99, 215)
betahat[1] + c(-1, 1)*CT*sqrt(VCM[1, 1])
betahat[2] + c(-1, 1)*CT*sqrt(VCM[2, 2])
betahat[3] + c(-1, 1)*CT*sqrt(VCM[3, 3])
# Or
confint(webe, level = 0.98)
```

------------------

Compute the mean response ($E(Y_h)$) for an apartment that had an area of 100 square meters and was 25 years old in 2005.

```{r}
Xh <- matrix(c(1, 100, 25), nrow = 1)
Yhath <- Xh %*% betahat
Yhath
# or
predict(webe, newdata = data.frame(area = 100, age = 25))
```
-----------

Find the standard error for the expected response given $X_h = (1, 100, 25)$.

```{r}
s2yhath <- Xh %*% VCM %*% t(Xh)
s2yhath
syhath <- sqrt(s2yhath)
syhath
```

---------

Construct a 90% confidence interval for the mean response given $X_h = (1, 100, 25)$.

```{r}
CT <- qt(.95, 215)
CT
Yhath + c(-1, 1)*CT*syhath
# Or 
predict(webe, newdata = data.frame(area = 100, age = 25), interval = "conf", level = 0.90)
```

----------

